Place of death in NRS COVID-19 deaths

### 2021-06-09 covid_nrs_place_of_death_visualisations.Rmd
- added deviation from mean graph showing deviations in absolute terms - added graph with both 2020 and 2021 trends on same week number scale, for easier comparison of same period both years; made 2020 transparent, but become more opaque later in the year when there isn't 2021 data yet, which I think is pretty neat!

## All of Scotland, deaths at home, for policy briefing

@@

## All of Scotland, deaths at home - date axis

@@

## All of Scotland, deaths at home - week number

```{r, fig.width=12, fig.height=9, warning=FALSE}
figure_deaths_at_home_wwek_number <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_vline(xintercept = 53.5, linetype = "dotted", colour="grey40", show.legend = FALSE) +  # line should be between last week 2020 and 1st week 2021
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  geom_text(aes(x = 10, y = 100, label = "Historic range (min & max)"), colour = "grey20") +
  geom_curve(aes(x = 10, xend = 14, y = 110, yend = 250), curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  geom_text(aes(x = 20, y = 50, label = "Historic mean"), colour = "grey20") +
  geom_curve(aes(x = 20, xend = 25, y = 60, yend = 270), curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  ## annotate year
  geom_text(aes(x=54, y=30, label = "2021"), hjust = 0, colour = "grey40") +
  NULL

## figure for a bit less than half A4 page is approx 14*11cm
save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_wwek_number"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home_wwek_number, width = 18, height = 15, units = "cm", dpi = 300)

figure_deaths_at_home_wwek_number
```

@@

@@

```{r}
(figure_deaths_by_cause <-
  merged_deaths_cause_of_death %>%
  filter(cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
   cause_of_death = fct_rev(cause_of_death)
 ) %>%
  group_by(place_of_death, cause_of_death) %>%
  summarise(
    deaths = sum(number_of_deaths),
    deaths_mean = sum(deaths_mean),
    .groups="drop"
  ) %>%
  ggplot() +
  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), alpha = 0.8, colour = "grey10") +
  geom_segment(aes(x = as.numeric(cause_of_death)-0.45, xend = as.numeric(cause_of_death)+0.45, y = deaths_mean, yend = deaths_mean), size = 0.8, colour = "grey1") +
  coord_flip() +
  scale_fill_viridis_d() +
  facet_wrap(~place_of_death) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  theme(
    legend.position = "none", 
    plot.caption = element_text(size = 10, colour = "gray60"),
    legend.title = element_blank(), 
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
    ) +
  labs(
    x = NULL,
    y = "Deaths",
    subtitle = paste0(c(paste0("Bars are deaths ",str_remove_all(date_range_text,"Data ")),"Line segments are mean annual deaths 2015-2019"),collapse="\n")
  ) +
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_deaths_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_by_cause, width = 10, height = 3, dpi = 300, units = "in")
```

>>>>

## Deaths at home, for policy briefing

@@

## Deaths at home - date axis

@@

### Home, deviation from mean

```{r}
(figure_deaths_at_home_deviation_from_mean <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>%
  mutate(
    deaths = deaths - deaths_mean,
    deaths_min = deaths_min - deaths_mean,
    deaths_max = deaths_max - deaths_mean
  ) %>%
  ggplot(aes(x = date_w_c)) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = 0), linetype = "dashed", colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = "Date w/c",
       y = "Weekly Deaths: difference from 2015-2019 average (N)",
       subtitle = paste0(c(date_range_text), collapse="\n"),
       caption = paste0(c(death_registrations_text,source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
   ## annotate historic range
  annotate(geom = "text", x=ymd("2021-01-07"), y=-120, label = "2021", hjust = 0, colour = "grey40") +
  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 16, year_number = 2020), y = -90, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 14, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 18, year_number = 2020), y = -80, yend = -50, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 37, year_number = 2020), y = -120, label = "Historic weekly average, set to 0", colour = "grey20") +
  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 37, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 42, year_number = 2020), y = -110, yend = -5, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  ## annotate year
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_deviation_from_mean"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_deviation_from_mean, width = 18, height = 14, dpi = 300, units = "cm")
```

## Deaths at home - week number

```{r, fig.width=12, fig.height=9, warning=FALSE}
figure_deaths_at_home_week_number <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_vline(xintercept = 53.5, linetype = "dotted", colour="grey40", show.legend = FALSE) +  # line should be between last week 2020 and 1st week 2021
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  geom_text(aes(x = 10, y = 100, label = "Historic range (min & max)"), colour = "grey20") +
  geom_curve(aes(x = 10, xend = 14, y = 110, yend = 250), curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  geom_text(aes(x = 20, y = 50, label = "Historic mean"), colour = "grey20") +
  geom_curve(aes(x = 20, xend = 25, y = 60, yend = 270), curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  ## annotate year
  geom_text(aes(x=54, y=30, label = "2021"), hjust = 0, colour = "grey40") +
  NULL

## figure for a bit less than half A4 page is approx 14*11cm
save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_week_number"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home_week_number, width = 18, height = 15, units = "cm", dpi = 300)

figure_deaths_at_home_week_number
```

@@

## 2020 and 2021 on same graph

For these graphs, we use the week number, and superimpose lines for 2020 and 2021!

### Home deaths, both years

```{r}
(figure_deaths_at_home_2020_and_2021 <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  select(-c(date_w_c, week_number_run_over, ref_period, ref_area)) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  # geom_line(aes(y = deaths, colour = deaths_type, alpha = year)) +
  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number), show.legend = FALSE) +
  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.8), guide = NULL) +
  annotate(geom = "text", x = 18, y = 550, label = "2020", alpha = 0.6) +
  annotate(geom = "text", x = 4, y = 530, label = "2021") +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x = 10, y = 100, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom = "curve", x = 10, xend = 14, y = 110, yend = 250, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x = 20, y = 50, label = "Historic mean", colour = "grey20") +
  annotate(geom = "curve", x = 20, xend = 25, y = 60, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020_and_2021"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2020_and_2021, width = 18, height = 14, dpi = 300, units = "cm")
```

@@

```{r}
(figure_deaths_by_cause <-
  merged_deaths_cause_of_death %>%
  filter(cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
  # filter(cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
   cause_of_death = fct_rev(cause_of_death)
 ) %>%
  group_by(place_of_death, cause_of_death) %>%
  summarise(
    deaths = sum(number_of_deaths),
    deaths_mean = sum(deaths_mean),
    .groups="drop"
  ) %>%
  ggplot() +
  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), alpha = 0.8, colour = "grey10") +
  geom_segment(aes(x = as.numeric(cause_of_death)-0.45, xend = as.numeric(cause_of_death)+0.45, y = deaths_mean, yend = deaths_mean), size = 0.8, colour = "grey1") +
  coord_flip() +
  scale_fill_viridis_d() +
  facet_wrap(~place_of_death) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  theme(
    legend.position = "none", 
    plot.caption = element_text(size = 10, colour = "gray60"),
    legend.title = element_blank(), 
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
    ) +
  labs(
    x = NULL,
    y = "Deaths",
    subtitle = paste0(c(paste0("Bars are deaths ",str_remove_all(date_range_text,"Data ")),"Line segments are mean annual deaths 2015-2019"),collapse="\n")
  ) +
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_deaths_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_by_cause, width = 10, height = 3, dpi = 300, units = "in")

(figure_deaths_by_cause_across_places <-
  merged_deaths_cause_of_death %>%
  filter(cause_of_death!="All" & place_of_death %in% c("All")) %>%
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
   cause_of_death = fct_rev(cause_of_death)
 ) %>%
  group_by(place_of_death, cause_of_death) %>%
  summarise(
    deaths = sum(number_of_deaths),
    deaths_mean = sum(deaths_mean),
    .groups="drop"
  ) %>%
  ggplot() +
  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), alpha = 0.8, colour = "grey10") +
  geom_segment(aes(x = as.numeric(cause_of_death)-0.45, xend = as.numeric(cause_of_death)+0.45, y = deaths_mean, yend = deaths_mean), size = 0.8, colour = "grey1") +
  coord_flip() +
  scale_fill_viridis_d() +
  facet_wrap(~place_of_death) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  theme(
    legend.position = "none", 
    plot.caption = element_text(size = 10, colour = "gray60"),
    legend.title = element_blank(), 
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
    ) +
  labs(
    x = NULL,
    y = "Deaths",
    subtitle = paste0(c(paste0("Bars are deaths ",str_remove_all(date_range_text,"Data ")),"Line segments are mean annual deaths 2015-2019"),collapse="\n")
  ) +
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_deaths_by_cause_all_places"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_by_cause_across_places, width = 7, height = 3, dpi = 300, units = "in")
```
